---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- badges: start -->

[![R-CMD-check](https://github.com/ncn-foreigners/jointCalib/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ncn-foreigners/jointCalib/actions/workflows/R-CMD-check.yaml)
[![CRAN
status](https://www.r-pkg.org/badges/version/jointCalib)](https://CRAN.R-project.org/package=jointCalib)
[![CRAN
downloads](http://cranlogs.r-pkg.org/badges/grand-total/jointCalib?color=blue)](https://cran.r-project.org/package=jointCalib )
<!-- badges: end -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview

## Details

A small package for joint calibration of totals and quantiles (see [Beręsewicz and Szymkowiak (2023)](https://arxiv.org/abs/2308.13281) working paper for details). The
package combines the following approaches:

-   Deville, J. C., and Särndal, C. E. (1992). [Calibration estimators
    in survey
    sampling](https://www.tandfonline.com/doi/abs/10.1080/01621459.1992.10475217).
    Journal of the American statistical Association, 87(418), 376-382.
-   Harms, T. and Duchesne, P. (2006). [On calibration estimation for
    quantiles](https://www150.statcan.gc.ca/n1/pub/12-001-x/2006001/article/9255-eng.pdf).
    Survey Methodology, 32(1), 37.
-   Wu, C. (2005) [Algorithms and R codes for the pseudo empirical
    likelihood method in survey
    sampling](https://www150.statcan.gc.ca/n1/pub/12-001-x/2005002/article/9051-eng.pdf),
    Survey Methodology, 31(2), 239.
-   Zhang, S., Han, P., and Wu, C. (2023) [Calibration Techniques
    Encompassing Survey Sampling, Missing Data Analysis and Causal
    Inference](https://onlinelibrary.wiley.com/doi/10.1111/insr.12518),
    International Statistical Review 91, 165--192.

which allows to calibrate weights to known (or estimated) totals and
quantiles jointly. As an backend for calibration
[sampling](https://CRAN.R-project.org/package=sampling)
(`sampling::calib`), [laeken](https://CRAN.R-project.org/package=laeken)
(`laeken::calibWeights`),
[survey](https://CRAN.R-project.org/package=survey) (`survey::grake`) or
[ebal](https://CRAN.R-project.org/package=ebal) (`ebal::eb`) package can
be used. One can also apply empirical likelihood using codes from Wu
(2005) with support of `stats::constrOptim` as used in Zhang, Han and Wu
(2022).

| backend    | method                                        | function called        |
|---------------|-------------------------------------------|---------------|
| `sampling` | `c("raking", "linear", "logit", "truncated")` | `sampling::calib`      |
| `laeken`   | `c("raking", "linear", "logit")`              | `laeken::calibWeights` |
| `survey`   | `c("raking", "linear", "logit", "sinh")`      | `survey::grake` |
| `ebal`     | `eb`                                          | `ebal::eb`             |
| `base`     | `el`                                         | R code and `stats::constrOptim` |

Currently supports:

-   calibration of quantiles,
-   calibration of quantiles and totals,
-   calibration using standard calibration, empirical likelihood and
    entropy balancing method.

Further plans:

-   generalized calibration via `sampling::gencalib`,
-   calibrated / covariate balancing propensity score (see [causal](https://github.com/ncn-foreigners/jointCalib/tree/causal) branch for initial development),
-   calibration for Gini and other metrics,
-   observational studies / causal inference (see [causal](https://github.com/ncn-foreigners/jointCalib/tree/causal) branch for initial development)
-   ...

## Funding

Work on this package is supported by the the National Science Centre,
OPUS 22 grant no. 2020/39/B/HS4/00941.

## Installation

You can install the development version of `jointCalib` from GitHub
with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ncn-foreigners/jointCalib")
```

## Examples

Load packages

```{r example1, message=FALSE}
library(jointCalib)
library(survey)
library(laeken)
library(ebal)
```

### Example 1 -- census case

Based on Haziza, D., and Lesage, É. (2016). A discussion of weighting
procedures for unit nonresponse. Journal of Official Statistics, 32(1),
129-145.

```{r example1-generate-data}
set.seed(20230817)
N <- 1000
x <- runif(N, 0, 80)
#y <- 1000+10*x+rnorm(N, 0, 300)
#y <- exp(-0.1 + 0.1*x) + rnorm(N, 0, 300)
#y <- rbinom(N, 1, prob = 1/(exp(-0.5*(x-55))+1))
y <- 1300-(x-40)^2 + rnorm(N, 0, 300)
#p <- rbinom(N, 1, prob = 0.2+0.6*(1 + exp(-5 + x/8))^-1)
p <- rbinom(N, 1, prob = 0.07+0.45*(x/40-1)^2+0.0025*x)
#p <- rbinom(N, 1, prob = (1.2+0.024*x)^-1)
#p <- rbinom(N, 1, prob = exp(-0.2 - 0.014*x))
probs <- seq(0.1, 0.9, 0.1)
quants_known <- list(x=quantile(x, probs))
totals_known <- c(x=sum(x))
df <- data.frame(x, y, p)
df_resp <- df[df$p == 1, ]
df_resp$d <- N/nrow(df_resp)
y_quant_true <- quantile(y, probs)
head(df_resp)
```

### Using `jointCalib` package

example 1a: calibrate only quantiles (deciles)

```{r example1-quantiles}
result1 <- joint_calib(formula_quantiles = ~x,
                      data=df_resp,
                      dweights=df_resp$d,
                      N = N,
                      pop_quantiles = quants_known,
                      method = "linear",
                      backend = "sampling")
result1
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result1$g*df_resp$d, probs))
```

example 1b: calibrate only quantiles (deciles)

```{r example1-joint}
result2 <- joint_calib(formula_totals = ~x,
                       formula_quantiles = ~x,
                       data = df_resp,
                       dweights = df_resp$d,
                       N = N,
                       pop_quantiles = quants_known,
                       pop_totals = totals_known,
                       method = "linear",
                       backend = "sampling")
summary(result2$g)
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result2$g*df_resp$d, probs))
```

We can restrict weights to specific range using `logit`.

```{r example1-joint-logit}
result3 <- joint_calib(formula_totals = ~x,
                       formula_quantiles = ~x,
                       data = df_resp,
                       dweights = df_resp$d,
                       N = N,
                       pop_quantiles = quants_known,
                       pop_totals = totals_known,
                       method = "logit",
                       backend = "sampling", 
                       maxit = 500,
                       bounds = c(0, 3))

summary(result3$g)
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result3$g*df_resp$d, probs))
```

Empirical likelihood method can be applied using the following code

```{r example1-joint-el}
result4a <- joint_calib(formula_quantiles = ~ x,
                        data = df_resp,
                        dweights = df_resp$d,
                        N = N,
                        pop_quantiles = quants_known,
                        method = "el",
                        backend = "base")
summary(result4a$g)

result4b <- joint_calib(formula_totals = ~ x,
                        formula_quantiles = ~ x,
                        data = df_resp,
                        dweights = df_resp$d,
                        N = N,
                        pop_quantiles = quants_known,
                        pop_totals = totals_known,
                        method = "el",
                        backend = "base")

summary(result4b$g)

result4c <- calib_el(X = result4b$Xs[, c(1, 11)],
                     d = df_resp$d, 
                     totals = c(N,totals_known),
                     maxit = 50,
                     tol = 1e-8)

summary(result4c)
```

```{r}
data.frame(true = quants_known$x, 
           est = weightedQuantile(df_resp$x, result2$g*df_resp$d, probs),
           est_el0 = weightedQuantile(df_resp$x, result4c*df_resp$d, probs),
           est_el1 = weightedQuantile(df_resp$x, result4a$g*df_resp$d, probs),
           est_el2 = weightedQuantile(df_resp$x, result4b$g*df_resp$d, probs))
```

Entropy balancing method can be applied using the following code

```{r example1-joint-eb}
result5 <- joint_calib(formula_quantiles = ~ x,
                        data = df_resp,
                        dweights = df_resp$d,
                        N = N,
                        pop_quantiles = quants_known,
                        method = "eb")
summary(result5$g)
```

Finally, compare all method with true Y distribution where

-   `est_y1` -- weights calibrated to quantiles only,
-   `est_y2` -- weights calibrated to quantiles and totals,
-   `est_y3` -- weights calibrated to quantiles and totals using `logit`
    distance and range limitations,
-   `est_y4` -- weights calibrated to means only using `el`.
-   `est_y5` -- weights calibrated to quantiles only using `el`.
-   `est_y6` -- weights calibrated to quantiles and totals using `el`.
-   `est_y7` -- weights calibrated to quantiles and totals using `eb`.

```{r results-combined}
results_y <- data.frame(true_y = y_quant_true,
                        est_y1 = weightedQuantile(df_resp$y, result1$g * df_resp$d, probs),
                        est_y2 = weightedQuantile(df_resp$y, result2$g * df_resp$d, probs),
                        est_y3 = weightedQuantile(df_resp$y, result3$g * df_resp$d, probs),
                        est_y4 = weightedQuantile(df_resp$y, result4c * df_resp$d, probs),
                        est_y5 = weightedQuantile(df_resp$y, result4a$g * df_resp$d, probs),
                        est_y6 = weightedQuantile(df_resp$y, result4b$g * df_resp$d, probs),
                        est_y7 = weightedQuantile(df_resp$y, result5$g * df_resp$d, probs))

results_y
```

Here is the sum of squares and calibration of totals and means seems to
be the best.

```{r}
apply(results_y[, -1], 2, FUN = function(x) sum((x-results_y[,1])^2))
```

### Using `survey` package

example 1a: calibrate only quantiles (deciles)

```{r example1-survey}
A <- joint_calib_create_matrix(X_q = model.matrix(~x+0, df_resp), N = N, pop_quantiles = quants_known)
colnames(A) <- paste0("quant_", gsub("\\D", "", names(quants_known$x)))
A <- as.data.frame(A)
df_resp <- cbind(df_resp, A)
head(df_resp)
m1 <- svydesign(ids = ~1, data = df_resp, weights = ~d)
quants_formula <- as.formula(paste("~", paste(colnames(A), collapse = "+")))
quants_formula
svytotal(quants_formula, m1)
```

Calibration using `calibrate`

```{r}
pop_totals <- c(N, probs)
names(pop_totals) <- c("(Intercept)", colnames(A))
m1_cal <- calibrate(m1, quants_formula, pop_totals)
svytotal(quants_formula, m1_cal)
```

Calibration using `grake` (low level)

```{r}
g1 <- grake(mm = as.matrix(cbind(1, A)),
            ww = df_resp$d,
            calfun = cal.linear,
            population = pop_totals,
            bounds = list(lower = -Inf, upper = Inf),
            epsilon = 1e-7,
            verbose = FALSE,
            maxit = 50,
            variance = NULL)
summary(as.numeric(g1))
```

## Example 2 -- non-probability samples

## Example 3 -- observational studies / causal inference
