---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# Overview

## Details

A small package for joint calibration of totals and quantiles. The package combines the following approaches:

+ Deville, J. C., and Särndal, C. E. (1992). [Calibration estimators in survey sampling](https://www.tandfonline.com/doi/abs/10.1080/01621459.1992.10475217). Journal of the American statistical Association, 87(418), 376-382.
+ Harms, T. and Duchesne, P. (2006). [On calibration estimation for quantiles](https://www150.statcan.gc.ca/n1/pub/12-001-x/2006001/article/9255-eng.pdf). Survey Methodology, 32(1), 37.

which allows to calibrate weights to known (or estimated) totals and quantiles jointly. As an backend for calibration [sampling](https://cran.r-project.org/web/packages/sampling) (`sampling::calib`) or [laeken](https://cran.r-project.org/web/packages/laeken) (`laeken::calibWeights`) package can be used.

Curently supports:

  + calibration of totals
  + calibration of quantiles

Further plans: 

  + calibration for Gini and other metrices
  + observational studies / causal inference
  + ...
  

## Funding

Work on this package is supported by the the National Science Center, OPUS 22 grant no. 2020/39/B/HS4/00941.


## Installation

You can install the development version of `jointCalib` from GitHub with:

```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ncn-foreigners/jointCalib")
```


## Examples 

### Example 1  

Based on Haziza, D., and Lesage, É. (2016). A discussion of weighting procedures for unit nonresponse. Journal of Official Statistics, 32(1), 129-145.

```{r example1, message=FALSE}
library(jointCalib)
library(survey)
library(laeken)
```


```{r example1-generate-data}
N <- 1000
x <- runif(N, 0, 80)
y <- exp(-0.1 + 0.1*x) + rnorm(N, 0, 300)
p <- rbinom(N, 1, prob = exp(-0.2 - 0.014*x))
probs <- seq(0.1, 0.9, 0.1)
quants_known <- list(x=quantile(x, probs))
totals_known <- c(x=sum(x))
df <- data.frame(x, y, p)
df_resp <- df[df$p == 1, ]
df_resp$d <- N/nrow(df_resp)
y_quant_true <- quantile(y, probs)
head(df_resp)
```

### Using `jointCalib` package

example 1a: calibrate only quantiles (deciles)

```{r example1-quantiles}
result1 <- joint_calib(formula_quantiles = ~x,
                      data=df_resp,
                      dweights=df_resp$d,
                      N = N,
                      pop_quantiles = quants_known,
                      method = "linear",
                      backend = "sampling")
summary(result1$g)
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result1$g*df_resp$d, probs))
```

example 1b: calibrate only quantiles (deciles)

```{r example1-joint}
result2 <- joint_calib(formula_totals = ~x,
                       formula_quantiles = ~x,
                       data = df_resp,
                       dweights = df_resp$d,
                       N = N,
                       pop_quantiles = quants_known,
                       pop_totals = totals_known,
                       method = "linear",
                       backend = "sampling")
summary(result2$g)
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result2$g*df_resp$d, probs))
```

We can restrict weights to specific range using `logit`.

```{r example1-joint-logit}
result3 <- joint_calib(formula_totals = ~x,
                       formula_quantiles = ~x,
                       data = df_resp,
                       dweights = df_resp$d,
                       N = N,
                       pop_quantiles = quants_known,
                       pop_totals = totals_known,
                       method = "logit",
                       backend = "sampling", 
                       maxit = 500,
                       bounds = c(0, 2))

summary(result3$g)
```

```{r}
data.frame(true = quants_known$x, est = weightedQuantile(df_resp$x, result3$g*df_resp$d, probs))
```

### Using `survey` package

example 1a: calibrate only quantiles (deciles)

```{r example1-survey}
A <- joint_calib_create_matrix(X_q = model.matrix(~x+0, df_resp), N = N, pop_quantiles = quants_known)
colnames(A) <- paste0("quant_", gsub("\\D", "", names(quants_known$x)))
A <- as.data.frame(A)
df_resp <- cbind(df_resp, A)
head(df_resp)
m1 <- svydesign(ids = ~1, data = df_resp, weights = ~d)
quants_formula <- as.formula(paste("~", paste(colnames(A), collapse = "+")))
quants_formula
svytotal(quants_formula, m1)
```

Calibration using `calibrate`

```{r}
pop_totals <- c(N, probs)
names(pop_totals) <- c("(Intercept)", colnames(A))
m1_cal <- calibrate(m1, quants_formula, pop_totals)
svytotal(quants_formula, m1_cal)
```
Calibration using `grake` (low level)

```{r}
g1 <- grake(mm = as.matrix(cbind(1, A)),
            ww = df_resp$d,
            calfun = cal.linear,
            population = pop_totals,
            bounds = list(lower = -Inf, upper = Inf),
            epsilon = 1e-7,
            verbose = FALSE,
            maxit = 50,
            variance = NULL)
summary(as.numeric(g1))
```

