---
title: "Balancing distributions for observational studies"
author: "Maciej Beręsewicz"
output: 
    html_vignette:
        df_print: kable
        toc: true
        number_sections: true
        fig_width: 6
        fig_height: 4
vignette: >
  %\VignetteIndexEntry{Balancing distributions for observational studies}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
link-citations: true
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Packages

Load packages for the example

```{r setup}
library(jointCalib)
library(CBPS)
library(ebal)
library(laeken)
```

Read the the `LaLonde` data.

```{r}
data("LaLonde")
head(LaLonde)
```

# ATT with entropy balancing and other methods

## Single variable: `age`

Consider balancing `age` variable.

```{r}
dens0 <- density(LaLonde$age[LaLonde$treat == 0])
dens1 <- density(LaLonde$age[LaLonde$treat == 1])
plot(dens0, main="Distribution of age", xlab="Age", ylim=c(0, max(dens0$y, dens1$y)))
lines(dens1, lty=2, col="blue")
legend("topright", legend=c("Control", "Treatment"), lty=c(1,2), col=c("black", "blue"))
```

Balancing using `ebal::ebalance` and `joint_calib_att` with method `eb` which uses `ebal` package as backend. For DEB we use deciles `probs = seq(0.1, 0.9, 0.1)` and balance mean as well (`formula_means = ~ age`). 

```{r}
bal_standard <- ebalance(LaLonde$treat, X = LaLonde[, "age"])
bal_quant <- joint_calib_att(formula_means = ~ age, 
                             formula_quantiles = ~ age, 
                             treatment =  ~ treat, 
                             data = LaLonde, 
                             probs = seq(0.1, 0.9, 0.1),
                             method = "eb")
bal_quant
```

Compare weighted distributions with treatment group distribution.

```{r}
dens0 <- density(LaLonde$age[LaLonde$treat == 0], weights = bal_standard$w/sum(bal_standard$w))
dens1 <- density(LaLonde$age[LaLonde$treat == 0], weights = bal_quant$g/sum(bal_quant$g))
dens2 <- density(LaLonde$age[LaLonde$treat == 1])
plot(dens0, main="Distribution of age", xlab="Age", ylim=c(0, max(dens0$y, dens1$y)))
lines(dens1, lty=2, col="blue")
lines(dens2, lty=3, col="red")
legend("topright", 
       legend=c("EB", "DEB", "Treatment"), 
       lty=c(1,2, 3), 
       col=c("black", "blue", "red"))
```

## More variables 

Now, consider three variables: `married`, `age` and `educ`.

```{r}
dens0 <- density(LaLonde$educ[LaLonde$treat == 0])
dens1 <- density(LaLonde$educ[LaLonde$treat == 1])
plot(dens0, main="Distribution of education", xlab="Education", ylim=c(0, max(dens0$y, dens1$y)))
lines(dens1, lty=2, col="blue")
legend("topleft", legend=c("Control", "Treatment"), lty=c(1,2), col=c("black", "blue"))
```

```{r}
bal_standard <- ebalance(LaLonde$treat, X = LaLonde[, c("married", "age", "educ")])
bal_quant <- joint_calib_att(formula_means = ~ married + age + educ, 
                             formula_quantiles = ~ age + educ, 
                             treatment =  ~ treat, 
                             data = LaLonde, 
                             method = "eb")
bal_quant
```

Compare education

```{r}
dens0 <- density(LaLonde$educ[LaLonde$treat == 0], weights = bal_standard$w/sum(bal_standard$w))
dens1 <- density(LaLonde$educ[LaLonde$treat == 0], weights = bal_quant$g/sum(bal_quant$g))
dens2 <- density(LaLonde$educ[LaLonde$treat == 1])
plot(dens0, main="Distribution of Education", xlab="Education", ylim=c(0, max(dens0$y, dens1$y)))
lines(dens1, lty=2, col="blue")
lines(dens2, lty=3, col="red")
legend("topleft", 
       legend=c("EB", "DEB", "Treatment"), 
       lty=c(1,2, 3), 
       col=c("black", "blue", "red"))
```

Compare age

```{r}
dens0 <- density(LaLonde$age[LaLonde$treat == 0], weights = bal_standard$w/sum(bal_standard$w))
dens1 <- density(LaLonde$age[LaLonde$treat == 0], weights = bal_quant$g/sum(bal_quant$g))
dens2 <- density(LaLonde$age[LaLonde$treat == 1])
plot(dens0, main="Distribution of age", xlab="Age", ylim=c(0, max(dens0$y, dens1$y)))
lines(dens1, lty=2, col="blue")
lines(dens2, lty=3, col="red")
legend("topright", 
       legend=c("EB", "DEB", "Treatment"), 
       lty=c(1,2, 3), 
       col=c("black", "blue", "red"))
```

Compare estimates of ATT using EB and DEB

```{r}
c(EB = with(LaLonde, mean(re74[treat == 1]) - weighted.mean(re74[treat == 0], bal_standard$w)),
  DEB = with(LaLonde, mean(re74[treat == 1]) - weighted.mean(re74[treat == 0], bal_quant$g)))
```

Compare QTT(0.5) using EB and DEB

```{r}
c(EB = with(LaLonde, median(re74[treat == 1]) - weightedMedian(re74[treat == 0], bal_standard$w)),
  DEB = with(LaLonde, median(re74[treat == 1]) - weightedMedian(re74[treat == 0], bal_quant$g)))
```

# ATT with CBPS and DPS

