---
title: "A minimal example of calibration for totals/means and quantiles in R, Python and Stata"
author: "Maciej Beręsewicz"
format: 
  html:
    self-contained: true
    table-of-contents: true
    number-sections: true
    df-print: kable
editor: source
execute: 
  eval: true
  warning: false
  message: false
toc-title: Contents
lang: en
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(engine.path = list(
  python = "/Library/Frameworks/Python.framework/Versions/3.10/bin/python3.10"
))
```

# Calibration for quantiles in R 

Minimal example using base R and `survey` package.

```{r}
library(survey)
```

Read the data (this is a csv file created based on `apiclus1` from `data(api)`).

```{r}
apiclus1 <- read.csv("apiclus1-example.csv")
head(apiclus1)
```

Say we calibrate to the following quantities

+ N = 6194
+ stypeH = 755
+ stypeM=1018
+ api99=3914069
+ api99_med = 650 (this is set arbitrary)

So we need to create new variabl `api99_med` 

```{r}
N <- 6194
api99_med_pop <- 670 ## arbitrary 
api99_sorted <- sort(apiclus1$api99) ## sort data and store in new vector
pos <- which(api99_sorted <= api99_med_pop) ## check with are lower or equal to pop median
n_pos <- NROW(pos) ## count cases
Q_L <- api99_sorted[n_pos] ## select lower 
Q_U <- api99_sorted[n_pos + 1] ## select upper
Q_B <- (api99_med_pop - Q_L) / (Q_U - Q_L) ## interpolation B
```

Add new variable to the dataset `apiclus1`.

```{r}
apiclus1$api99_med <- numeric(nrow(apiclus1)) ## create vector with 0
apiclus1$api99_med[apiclus1$api99 < Q_L] <- 1/N 
apiclus1$api99_med[apiclus1$api99 == Q_U] <- Q_B/N
head(apiclus1[, c("api99", "api99_med")])
```

Calibration using `survey::calibrate`.

```{r}
pop.totals<-c(`(Intercept)`=N, stypeH=755, stypeM=1018, api99=3914069, api99_med=0.50) ## vector of totals
pop.totals
```

Declare design and calibrate
```{r}
dclus1<-svydesign(id=~dnum, weights=~pw, data=apiclus1, fpc=~fpc) ## definition of design
dclus1g3 <- calibrate(dclus1,  formula = ~stype+api99+api99_med,  population = pop.totals) ## calibration
```

Compare results

```{r}
c(before = svyquantile(~api99, dclus1, quantile = 0.5)$api99[1],
  after = svyquantile(~api99, dclus1g3, quantile = 0.5)$api99[1])
```

Check `api99_med` variable if it sums up to 0.5

```{r}
c(before = svytotal(~api99_med, dclus1), after = svytotal(~api99_med, dclus1g3))
```

Note that in small sample sizes quantiles may not be reproduced excatly due to rounding


# Calibraion for quantiles in Python using `pandas` and `samplics`

Under `samplics-0.4.8`

```{bash, eval = FALSE} 
pip install samplics
pip install statsmodels
pip install np
pip install pandas
```

```{python}
from statsmodels.stats.weightstats import DescrStatsW
import numpy as np
import pandas as pd
from samplics.weighting import SampleWeight
```

```{python}
apiclus1 = pd.read_csv("apiclus1-example.csv")
apiclus1["pop"] = 1
apiclus1["stypeH"] = 0
apiclus1["stypeM"] = 0
apiclus1.loc[apiclus1['stype'] == 'H', 'stypeH'] = 1
apiclus1.loc[apiclus1['stype'] == 'M', 'stypeM'] = 1
apiclus1.head()
```

```{python}
N = 6194
api99_med_pop = 670
api99_sorted = np.sort(apiclus1['api99'])
pos = np.where(api99_sorted <= api99_med_pop)
n_pos = pos[0].shape[0]
Q_L = api99_sorted[n_pos]
Q_U = api99_sorted[n_pos + 1]
Q_B = (api99_med_pop - Q_L) / (Q_U - Q_L)
apiclus1['api99_med'] = 0
apiclus1.loc[apiclus1['api99'] < Q_L, 'api99_med'] = 1/N 
apiclus1.loc[apiclus1['api99'] == Q_U, 'api99_med'] = Q_B/N
apiclus1[["api99", "api99_med"]].head()
```

Declare population totals

```{python}
pop_totals = {"pop": 6194, "stypeH" : 755, "stypeM" : 1018, "api99": 3914069, "api99_med": 0.5}
pop_totals
```

Calibration using `samplics` (linear function)

```{python}
apiclus1["w_cal"] = SampleWeight().calibrate(
    apiclus1["pw"], apiclus1[["pop", "stypeH", "stypeM", "api99", "api99_med"]], 
    pop_totals)
```

Weighted quantiles are not implemented so we use `DescrStatsW` from `statsmodels.stats.weightstats`.

```{python}
wq1 = DescrStatsW(data=apiclus1["api99"], weights=apiclus1["pw"])
wq2 = DescrStatsW(data=apiclus1["api99"], weights=apiclus1["w_cal"])

{"Before": wq1.quantile(probs=0.5, return_pandas=False)[0],
 "After": wq2.quantile(probs=0.5, return_pandas=False)[0]}
```

Check `api99_med` variable if it sums up to 0.5

```{python}
wq1 = DescrStatsW(data=apiclus1["api99_med"], weights=apiclus1["pw"])
wq2 = DescrStatsW(data=apiclus1["api99_med"], weights=apiclus1["w_cal"])

{"Before": wq1.sum, "After": wq2.sum}
```

# Calibraion for quantiles in Stata using `svyset` (`regress`)

```{stata, eval = FALSE}
import delimited using "your_file.csv", clear
set more off

* Define the constants
global N = 6194
global api99_med_pop = 650

* Create a copy of api99 as a temporary variable
gen api99_copy = api99

* Sort the temporary variable
sort api99_copy

* Define new variable pos to store the values that are lower or equal to api99_med_pop
gen pos = (api99_copy <= $api99_med_pop)

* Count the number of cases where pos is true
egen n_pos = total(pos)

* Define Q_L and Q_U based on n_pos
gen Q_L = api99_copy if _n == n_pos
gen Q_U = api99_copy if _n == n_pos + 1

* Replace missing Q_L and Q_U with respective preceding and following values 
replace Q_L = Q_L[_n-1] if _n > 1 & missing(Q_L)
replace Q_U = Q_U[_n+1] if _n < _N & missing(Q_U)

* Interpolation B
gen Q_B = ($api99_med_pop - Q_L) / (Q_U - Q_L)

* Create a new variable api99_med with 0
gen api99_med = 0

* Update api99_med according to conditions
replace api99_med = 1/$N if api99 < Q_L
replace api99_med = Q_B/$N if api99 == Q_U

* Drop the temporary variables
drop api99_copy Q_L Q_U pos n_pos Q_B

* Show the first few rows
list in 1/5

```

